Data Specialization for a Staged Lambda Calculus

### 2015-02-24 paper/example.tex
small edits all over

We define a staged version of quickselect in \ref{fig:qs-staged}, writing
first-stage computations in red, and second-stage computations in blue. While
\texttt{qSelect} sends a \textrm{list} and $\rmint$ to an $\rmint$,
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a \textrm{list} (ignoring the $\curr$ for the moment) and a $\fut\rmint$
(an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

@@

Our type system, discussed in \ref{sec:semantics}, ensures that the staging
annotations are consistent, in the sense that computations marked as
first-stage cannot depend on ones marked as second-stage. The $\curr$ and
\texttt{grnd} annotations further distinguish those first-stage computations
which do not contain any embedded second-stage code. (Code \emph{inside}
\texttt{grnd}, like the \texttt{part}ition function, is guaranteed not to
contain second-stage code; other code, like \texttt{qsStaged}, may.) We will
discuss the importance of this additional distinction in \ref{sec:splitting}.

@@

An astute programmer, having noticed that \texttt{quickselect} can be staged in
this fashion, might try to split it into a pair of functions, one which performs
all the work depending only on \texttt{l} (the first stage), and one which uses
\texttt{k} and the result of that work on \texttt{l} to compute the element with
rank \texttt{k} in \texttt{l}.

Intuitively, \texttt{l} determines the result of all calls to
\texttt{part}ition, and \texttt{k} only determines which calls are made. So we
can preprocess \texttt{l} by recursively dividing it into halves smaller and
greater than the pivot---that is, building a binary search tree. Then, once we
have \texttt{k}, we can recur on this tree, choosing whichever branch has the
\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. And because
\texttt{part}ition contains no second-stage code, we can run it entirely in the
first stage!

@@

Assuming \texttt{l} contains $n$ elements, this optimization changes the
asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$ to
$\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.

>>>>

We define a staged version of quickselect in \ref{fig:qs-staged}, writing
first-stage computations in red, and second-stage computations in blue. While
\texttt{qSelect} sends a \textrm{list} and $\rmint$ to an $\rmint$,
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$
(an $\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

@@

\lang's type system, discussed in \ref{sec:semantics}, ensures that the staging
annotations are consistent, in the sense that computations marked as
first-stage cannot depend on ones marked as second-stage. The $\curr$ and
\texttt{grnd} (``ground'') annotations further distinguish those first-stage computations
which do not contain any embedded second-stage code. (Code \emph{inside}
\texttt{grnd}, like the \texttt{part}ition function, is guaranteed not to
contain second-stage code; first stage code outside a \texttt{grnd}, 
like \texttt{qsStaged}, may.) We will
discuss the importance of this additional distinction in \ref{sec:splitting}.

@@

An astute programmer, having noticed that \texttt{quickselect} can be staged in
this fashion, might try to split it into a pair of functions, one which performs
all the work depending only on \texttt{l} (the first stage), and one which uses
that partial result and \texttt{k} to compute the element with
rank \texttt{k} in \texttt{l}.

Intuitively, \texttt{l} determines the result of all calls to
\texttt{part}ition, and \texttt{k} only determines which calls are made. So we
can preprocess \texttt{l} by recursively dividing it into halves smaller and
greater than the pivot---that is, building a binary search tree. Then, once we
have \texttt{k}, we can recur on this tree, choosing whichever branch has the
\texttt{k}${}^\textit{th}$ leftmost element until we reach a leaf. And because
\texttt{part}ition contains no second-stage code, we can run it entirely in the
first stage.

@@

Assuming \texttt{l} contains $n$ elements, this optimization changes the
asymptotic complexity from expected (randomized) $\Theta(n \cdot m)$ to
$\Theta(n\log{n} + m\log{n})$, which for any $m \approx n$ reduces the
complexity from $\Theta (n^2)$ to $\Theta(n\log{n})$---a significant improvement.

### 2015-02-24 paper/semantics.tex
small edits all over

Although our products and functions are restricted to types at the same world,
$\fut$ essentially allows us to build ``mixed-stage'' products and functions.
For example, quickselect is a function at world \bbonem\ which takes a
$\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
second-stage computation of an integer) to a $\fut\rmint$ (a second-stage
computation of an integer).

@@

We evaluate \lang\ terms in two stages. The first stage of evaluation:
\begin{enumerate}
\item fully reduces terms in \bbonep, as they are purely first-stage;
\item reduces terms in \bbonem, skipping over any world \bbtwo\ terms they may
contain (inside $\next$);
\item does not reduce terms in \bbtwo, but scans them for any \bbonem\ terms
they may contain (inside $\prev$) and reduces those.
\end{enumerate}
These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
big-step judgments, respectively (defined in \ref{fig:diaSemantics}).

This process ultimately produces a monostaged term, called a \emph{residual}
(\ref{fig:values}), all of whose redexes are second-stage computations. The
second stage of evaluation is therefore just ordinary (unstaged) evaluation of
this residual, which we notate $\redsym$. (For simplicity, we treat residuals
and \bbonep\ terms as living in a single language equipped with a standard
evaluation $\redsym$, whose definition we omit for space reasons.)

@@

For all constructs other than $\prev{e}$, $\redtwosym$ evaluation proceeds
recursively on all subterms, yielding a monostaged residual $q$. At $\prev{e}$,
we evaluate $e$ using $\redonesym$ to $[\xi;v]$. Because $\colmix e{\fut A}$,
$v$ must be $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
$\xi$ into a sequence of second-stage \texttt{let} bindings around $\hat{y}$,
using the auxiliary judgment $\reify\xi q{q'}$ defined in
\ref{fig:diaSemantics}, then return the resulting residual $q'$.

>>>>

Although products and functions are restricted to types at the same world,
$\fut$ allows costruction of ``mixed-stage'' products and functions.
For example, quickselect is a function at world \bbonem\ which takes a
$\curr{\rm list} \times \fut\rmint$ (a purely-first-stage list and a
second-stage integer) to a $\fut\rmint$ (a second-stage
computation of an integer).

@@

We evaluate \lang\ terms in two stages. The first stage of evaluation has 
three interdependent modes of operation, depending on the world of the term it is evaluating.
Specifically, first stage evaluation:
\begin{enumerate}
\item fully reduces terms in \bbonep, as they are purely first-stage;
\item reduces terms in \bbonem, save for any world \bbtwo\ terms they may
contain (inside \texttt{next}s);
\item scans terms in \bbtwo\ for any world \bbonem\ subterms (inside \texttt{prev}s), 
and reduces those in place.
\end{enumerate}
These tasks are accomplished by the $\redsym$, $\redonesym$, and $\redtwosym$
big-step judgments, respectively (defined in \ref{fig:diaSemantics}).

First-stage evaluation ultimately produces a monostaged term, called a \emph{residual}
(\ref{fig:values}), all of whose redexes are second-stage computations. The
second stage of evaluation is therefore just ordinary (unstaged) evaluation of
this residual, which we notate $\redsym$. (For simplicity, we treat residuals
and \bbonep\ terms as living in a single language equipped with a standard
evaluation $\redsym$, whose definition we omit for space reasons.)

@@

For all constructs other than $\prev{e}$, $\redtwosym$ evaluation proceeds
recursively on all subterms, yielding a monostaged residual $q$. At $\prev{e}$,
we evaluate $e$ using $\redonesym$ to $[\xi;v]$. Because $\colmix e{\fut A}$,
$v$ must have the canonical form $\next{\hat{y}}$ for some $\hat{y}$ bound in $\xi$. We \emph{reify}
$\xi$ into a sequence of second-stage \texttt{let} bindings around $\hat{y}$,
using the auxiliary judgment $\reify\xi q{q'}$ defined in
\ref{fig:diaSemantics}, then return the resulting residual $q'$.

### 2015-02-24 paper/splitting.tex
small edits all over

Although the type system of \lang\ ensures that we can perform all first-stage
computations before any second-stage ones, actually evaluating terms in this
fashion is somewhat intricate, because stages are syntactically interleaved
within terms. In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of 
syntactically % I don't like this word but temporarily here for continuity
separate programs corresponding to the two stages of computation in that term.

@@

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
we also provide a way to split those as well.
This is given by the new relation $e \splitonesym \mathcal{P}$,
where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computation that was in $e$, 
and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
second-stage computation that was in $e$.

Similar to above, we also define a $\sepredtwosym$ relation which can be used to evaluate \texttt{sepM} terms,
with one complication arising from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, consider the term
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
which reduces (via $\redonesym$) to the residual table and partial value
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}
Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.

We resolve this mismatch by introducing one more translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
which is just a technical device to convert residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
As before, we enforce the ``syntactically separated'' property structurally,
saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
$i$ is a monostage value representing the first-stage part of $v$,
and $q$ is a monostage term representing all of the second-stage computation in $\xi$ and $v$.
Note that masking exists only for the purposes of stating correctness; it is not part of the splitting algorithm itself.

@@

\subsection {Three World System?}

Why do we need a three world system?

>>>>

Although the type system of \lang\ ensures that we can perform all first-stage
computations before any second-stage ones, actually evaluating terms in this
fashion is rather involved, because stages are syntactically interleaved
within terms. In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of 
syntactically % I don't like this word but temporarily here for continuity
separate programs corresponding to the two stages of computation in that term.

@@

Since terms at world \bbtwo\ can depend on terms at \bbonem\ (via \texttt{prev}),
we must also provide a way to split those as well.
This is given by the new relation $e \splitonesym \mathcal{P}$,
where $\mathcal{P}$ must have the form $\pipeM c l r$ where  
$c$ (called the {\em combined term}) is a monostage term encoding all the first-stage computation that was in $e$, 
and $l.r$ (again called the {\em resumer}) is a monostage term encoding all the
second-stage computation that was in $e$.

Similar to above, we also define a $\sepredtwosym$ relation 
which can be used to evaluate these seperated world \bbonep terms.
One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
For example, consider the term
\begin{lstlisting}
(next {1+2}, grnd{3+4})
\end{lstlisting}
which reduces (via $\redonesym$) to the residual table and partial value
\begin{lstlisting}
[yhat |-> 1+2] (next {yhat}, grnd{7})
\end{lstlisting}
Since this output has interleaved stages, theres no way a syntactically separated term could directly reduce to it.

We resolve this mismatch by introducing a translation, called {\em masking} and written $[\xi;v] \vsplito \mathcal{V}$,
which converts residual tables and partial values---like the those above---into a syntactically separated version $\mathcal{V}$.
As before, we enforce the ``syntactically separated'' property structurally,
saying that $\mathcal{V}$ must have the form $\mval i q$ (called a {\em masked value}) where 
$i$ is a monostage value representing the first-stage part of $v$,
and $q$ is a monostage term representing all of the second-stage computation in $\xi$ and $v$.
Note that masking is a technical device for the purpose of stating correctness of splitting at \bbonem; 
it is not part of the splitting algorithm itself.

@@

\subsection {Necessity of \bbonep}

We have not yet given a justification for why first stage code must be
partitioned between the worlds \bbonem\ and \bbonep.
Certainly one could imagine a simpler system without all the \texttt{grnd} annotations.  
Why wouldn't this work?

In order to be correct, any code at \bbonem\ must be split with the 
pessimistic assumption that it may result in work at the second stage.
When this assumption turns out not to be true---that is, purely monostage code---
then splitting may produce second stage code which is needlessly costly 
(an example of this will occur in the next section).
Detecting and optimizing this case is in general a global program analysis,
since functions can be passed around as values.  
So instead of relying on hefty analysis, 
we take the approach of adding enough structure to the input language's type system to
allow the input code to {\em prove} itself to be---in some parts---monostage.

We find that giving input with \texttt{grnd} annotations is not especially cumbersome in practice.
Yet even if a language designer desired a different implicit/explicit trade-off,
then the three-world \lang\ would still be useful as a typed intermediate representation.
